\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-02-07 paper/main.tex
done methods

\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger
graph convolutional model, based on the protein-protein interaction
(PPI) graph. The PPI dataset is represented by a graph $G=(V,E)$,
where each protein is represented by a vertex $v\in V$, and each edge
$e\in E\subseteq V\times V$ symbolizes an interaction between two
proteins. Additionally, we introduce mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v = x(v)$, where $d$ denotes the dimensionality of the node
features.

As described before, graph convolution has shown significant
performance increase in a variety of tasks. While there are various
methods out there we will only introduce the most basic one here. A
graph convolutional layer \citet{GCNConv} consists of a
learnable weight matrix followed by an aggregation step, formalized by
\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}
where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the
adjacency matrix with added self-loops for each vertex, $D$ is
described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal
matrix displaying the degree of each node, and $\Theta$ denotes the
learnable weight matrix. Added self-loops enforce that each node
representation is directly dependent on its own preceding
one. Notably, the number of graph convolutional layers stacked equals
the radius of relevant nodes for each vertex within the graph.

@@

While in this initial formulation the node-wise update step is defined
by the sum over all neighbouring node representations, we are able to
alter this formulation to another message passing scheme.  We are able
to rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$ and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we will generally only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We will denote
this generalized layer type as \textsc{GENConv}, following the
notation of \citet{PytorchGeometric}.  While the reordering is mainly
import for numerical stability, this alteration also addresses the
vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale, highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously, and the resulting information\todo{Which information?
  Specify} hold no information on their own connectivity.

@@

We employ another mechanism to avoid redundancy and fading signals in
stacked graph convolutional networks, which was introduced by
\citet{DeepGCN2019} and refined in \citet{DeeperGCN2020}. The authors
propose a residual network architecture and normalization scheme. This
structure is depicted in Figure~\ref{fig:ResGraphConvBlocks} resulting
in reusable residual blocks, which can be stacked multiple times,
thereby not losing focus of each node neighbourhood.

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparsely given with respect to the number of both drugs and proteins considered, the resulting training, validation and testing datasets are highly
imbalanced. As there are only $22.336$ links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins as justified in the results section. Furthermore, we used
early stopping to detect plateaus in the training process.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
w.r.t. previously described metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. \todo[inline]{Specify range of
  values searched here; even better if you have the intermediate
  results and can put them here (better: in the supplement).}

@@

To assess each model, we compute a variety of common metrics for binary classification. As the datasets are highly imbalanced, we chose the \textit{area under receiver operating characteristic}-score (AUROC) on training, validation and testing split. Therefor, we calculated the true positive rate (TPR), false positive rate (FPR) and precision score. We will further compute true positives (TP), false positives (FP), false negatives (FN) and finally true negatives (TN). 
We calculate the ROC by plotting the TPR against the FPR for various threshold settings for the model predictions. We eventually calculate the area under ROC curve utilizing trapezoidal approximations. We will refer to this measure as $MacroAUC$.

In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Note that drugs and proteins can be interchanged in this formulation, which we will denote with $MicroAUC_p$ and $MicroAUC_d$, respectively. Additionally, the $MicroAUC$ score may not be defined as, in some datasets some targets or drugs, respectively, do not have any interactions, leading to an infeasible $TPR$ and an undefined $AUROC$ score for that entity. For those entities we impute the $MicroAUC$ linearly, by using the accuracy for this subset.

>>>>

\subsubsection{Graph convolutional layers}

We included these molecular and ontology-based sub-models with a
larger graph convolutional model. The graph underlying the graph
convolutional neural network is based on the protein--protein
interaction (PPI) graph. The PPI dataset is represented by a graph
$G=(V,E)$, where each protein is represented by a vertex $v\in V$, and
each edge $e\in E\subseteq V\times V$ symbolizes an interaction
between two proteins. Additionally, we introduce a mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v := x(v)$, where $d$ denotes the dimensionality of the node
features.

% As described before, graph convolution has shown significant
% performance increase in a variety of tasks. While there are various
% methods out there we will only introduce the most basic one here. 
A graph convolutional layer \citet{GCNConv} consists of a learnable
weight matrix followed by an aggregation step, formalized by
\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}
where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the
adjacency matrix with added self-loops for each vertex, $D$ is
described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal
matrix displaying the degree of each node, and $\Theta$ denotes the
learnable weight matrix. Added self-loops enforce that each node
representation is directly dependent on its own preceding one. The
number of graph convolutional layers stacked equals the radius of
relevant nodes for each vertex within the graph.

@@

While in this initial formulation the node-wise update step is defined
by the sum over all neighbouring node representations, we are able to
alter this formulation to another message passing scheme.  We are able
to rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$ and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we will generally only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We will denote
this generalized layer type as \textsc{GENConv}, following the
notation of PyTorch Geometric \cite{PytorchGeometric}.  While the
reordering is mainly import for numerical stability, this alteration
also addresses the vanishing gradient problem for deeper convolutional
networks \cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale, highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.
The residual blocks are reusable and can be stacked multiple times. %, thereby not losing focus of each node neighbourhood.
\todo[inline]{Needs a clearer outline/description/figure of YOUR model
structure.}
The structure is depicted in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to
binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for a given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins. Furthermore, we used early stopping to detect plateaus in
the training process.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
with respect to our evaluation metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. \todo[inline]{Specify range of
  values searched here; even better if you have the intermediate
  results and can put them here (better: in the supplement).}

@@

To assess each model, we compute a variety of common metrics for
binary classification. As the datasets are highly imbalanced, we use
the area under the receiver operating characteristic curve
(AUROC) on training, validation and testing split. % We
% calculated the true positive rate (TPR), false positive rate (FPR), and
% precision score. We will further compute true positives (TP), false
% positives (FP), false negatives (FN) and finally true negatives (TN).

\todo[inline]{Needs a statement on how the PAIRS are ranked here and
  TPR/FPR is calculated based on the ranks of the DT-PAIRS.}
We calculate the AUROC by computing true positive rate at various
false positive rate thresholds and using trapezoidal approximations to
estimate the area under the curve. We refer to this measure as
$MacroAUC$.

\todo[inline]{For RH: check this again later:}
In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Drugs and proteins can be interchanged in this formulation, and we
refer to the different measures as a protein-centric microAUC
($MicroAUC_p$) and a drug-centric microAUC
($MicroAUC_d$). Additionally, the $MicroAUC$ score may not be defined
as, in some datasets some targets or drugs, respectively, do not have
any interactions, leading to an infeasible $TPR$ and an undefined
$AUROC$ score for that entity.
\todo[inline]{Don't understand this, can you add the formula:} For
those entities we impute the $MicroAUC$ linearly, by using the
accuracy for this subset.
